Headgroup structure and cation binding in phosphatidylserine lipid bilayers

### 2019-05-23 Manuscript/manuscriptPS.tex
Changes to affiliations, tiny modifications to text and typo corrections

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as specific binding; in the presence of calcium ions, PS can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain not well understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
We then use this data set
to assess,
in a wide range of available molecular dynamics (MD) simulation models (force fields) of PS,
the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes.
We find that none of the MD models reproduce the NMR data within experimental accuracy.
The best MD models, however, suggest that the carboxyl group in the PS headgroup
does not rotate freely. In line with our previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity; moreover, the response of PS headgroups to bound ions can  differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes. In red blood cells, for example,
PS lipids compose 8.5\% of the total lipid weight. 
The abundance, however, varies between different organelles, and up to
25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
PS lipids are vastly important biomolecules that interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while other protein sites attract PS lipids by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, deciphering the structural details
of lipid headgroups and the details of cation binding
are crucial for understanding the PS-mediated processes in the cell membranes.

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
While most monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroups with calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}. Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
That said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
to negatively charged and zwiterionic phospholipids, and that
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.

@@

  Notation 2$\times$[time] indicates that two independent MD runs was conducted.
  JC refers to the Joung-Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.

@@

lipid/counter-ions  & force field & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline

@@

The molecular level interpretation of these observations is,
however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}. Using this approach,
it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, that the cation binding to PC
lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

Here, we extend the available set of experimentally measured PS lipid headgroup and
glycerol backbone C--H bond order parameters
by measuring the signs of the order parameters using S-DROSS solid-state NMR spectroscopy.
The quality of headgroup structures and the ion binding affinity in
%MD simulations of lipid bilayers containing PS lipids
the available MD simulation models of PS lipids are then asessed based on the collected experimental data.
%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
The results pave the way
for development of lipid models that correctly describe 
the headgroup region of negatively charged lipids in physiological salt
conditions. Such force fields are expected to be useful in understanding
biological function of lipid headgroups and glycerol backbone, as
these are known to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.

@@

  Notation 2$\times$[time] indicates that two independent MD runs was conducted.

@@

lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline

@@

The headgroup and glycerol backbone C--H bond order parameter magnitudes of 1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine (POPS)
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The following experimental setups were used.

{\emph{C--H bond order parameters from the R-PDLF experiment.}} The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=$\,1.94\,ms and $\tau_2=$\,0.97\,ms.
The used radio frequency pulses had the following nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
50 kHz (SPINAL64 $^1$H decoupling pulses).
The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$, and 32 points in the indirect
dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral width of 149.5 ppm.

\emph{Order parameter signs from the S-DROSS experiment.}
The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. The used radio frequency pulses had the nutation
frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H\,SPINAL64 decoupling).
The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s, and a total of 8 points along $t_1$ were
measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral width of 149.5 ppm.

@@

\emph{Sample preparation.}
The sample was prepared simply by mixing POPS powder (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis can be conducted using a
Python program ({\tt calcOrderParameters.py}, available in Ref. \citenum{MATCHgit}) that utilizes the
MDAnalysis library \cite{agrawal11,gowers16}.
For united atom models, the positions of hydrogens were generated before the order parameter calculation using {\tt protonate} tool
of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup order parameters at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup order parameters
of negatively charged PS and PG lipids also exhibit systematic, but less
understood dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup order parameters from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

@@

Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
the concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16},
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  We can do this just before submission when citation numbers will not change anymore.}
\end{figure}

@@

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC})
and more accurate force fields are needed
to correctly describe the PC-PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive

>>>>

\author{Hanne Antila}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{NMR group - Institut for Physics, Martin-Luther University Halle-Wittenberg}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology, Dolgoprudny, Russia}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\affiliation{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{Markus S. Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry of the 
Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2, CZ-16610 Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas J. Piggot}
\affiliation{Chemistry, University of Southampton, Highfield, Southampton SO17 1BJ, U.K}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is a negatively
charged lipid commonly found in eukaryotic membranes, where it interacts with proteins via
nonspecific electrostatic interactions as well as via specific binding. Moreover, in the presence of calcium ions, PS can induce
membrane fusion and phase separation.
Molecular details of these phenomena remain poorly understood, partly
because accurate models to interpret the experimental data have not
been available. Here we
gather a set of experimental NMR data of C--H bond order parameters, $S_\mathrm{CH}$, for pure
PS and mixed PS:PC (phosphatidylcholine) lipid bilayers, 
and augment this data set by measuring with solid-state NMR S-DROSS the signs of $S_\mathrm{CH}$ in the POPS headgroup.
We then use this data set
to assess,
in a wide range of available molecular dynamics (MD) simulation models (force fields) of PS,
the accuracy of 1) the PS headgroup structures and 2) the ion binding to PS-containing membranes.
We find that none of the MD models reproduce the NMR data within experimental accuracy.
The best MD models, however, suggest that the carboxyl group in the PS headgroup
does not rotate freely. In line with our previous results for PC lipids,
none of the PS force fields correctly captures the cation binding affinity. Moreover, the response of PS headgroups to bound ions can differ from experiments even \emph{qualitatively}. The collected experimental dataset and simulation
results pave the way for the improvement of lipid force fields that correctly
describe the biologically relevant negatively charged membranes and their interactions with ions.
This work is part of the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes. In red blood cells, for example,
PS lipids compose 8.5\% of the total lipid weight. 
The abundance, however, varies between different organelles, and up to
25--35\% of the cytosolic leaflet of plasma membranes 
\cite{lemmon08,leventis10,li14} consists of PS lipids.
PS lipids are vastly important biomolecules that interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while other protein sites attract PS lipids by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, deciphering the structural details
of lipid headgroups and the details of cation binding
are crucial for understanding PS-mediated processes on cellular membranes.

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
owing to electrostatic interactions or the formation of hydrogen bonding network between the headgroups~\cite{browning80,buldt81}.
While most monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroups with calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}. Mixing PS lipids with PC lipids reduces their propensity to form strong complexes with multivalent ions and makes the PS headgroup less rigid~\cite{browning80,buldt81,roux90,roux91}.
That being said, some studies suggest Ca$^{2+}$ has similar specific binding affinity 
to negatively charged and zwiterionic phospholipids, and that
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the vicinity of the membrane~\cite{seelig90,sinn06}.

@@

  Notation 2$\times$[time] indicates that two independent MD runs were conducted.
  JC refers to the Joung--Cheatham ion parameters~\cite{joung08} and ff99 to the default Amber ion parameters~\cite{aqvist90}.

@@

lipid/counter-ions  & force field & \footnote{Number of lipid molecules with largest mole fraction}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files & \tabularnewline

@@

The molecular level interpretation of these observations is currently lacking, and classical molecular dynamics (MD) simulations have been widely used in efforts 
to understand the PS headgroup structure, its influence on lipid bilayer properties, and its
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}. In contrast,
the results from the same lipid model without the NBfix ion parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, indicate a significant binding affinity also to the phosphate region \cite{hallock18}.
Meanwhile, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest substantial
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that the lipid C--H bond order parameters, $S_\mathrm{CH}$,
can be utilized to resolve such controversies~\cite{botan15,catte16}. The $S_\mathrm{CH}$ can be
measured from NMR experiments with high accuracy and directly compared to simulations
in order to evaluate the quality of the simulation model, or to interpret the experiments \cite{ollila16}. Using this approach,
it has been established that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, that the cation binding to PC
lipid bilayers is overestimated~\cite{catte16}, and that the splitting of the order parameters at
carbon 2 in {\it sn}-2 lipid tails is correctly reproduced only in the CHARMM36 force field~\cite{piggot17}.% [https://pubs.acs.org/doi/abs/10.1021/acs.jctc.7b00643].
%
%Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
%\todo{Should we leave the mention of ECC out from the Introduction
%(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

Here, we extend the available set of experimentally measured PS lipid headgroup and
glycerol backbone C--H bond order parameters
by measuring the signs of these order parameters using S-DROSS solid-state NMR spectroscopy.
The quality of headgroup structures and the ion binding affinity in
%MD simulations of lipid bilayers containing PS lipids
the available MD simulation models of PS lipids are then assessed based on the collected experimental data.
%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
The results pave the way
for the development of lipid models that correctly describe 
the headgroup region of negatively charged lipids under physiological salt
conditions. Such force fields are expected to be extremely useful in understanding
biological function of lipid headgroups and glycerol backbone, as
these are known to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.

@@

  Notation 2$\times$[time] indicates that two independent MD runs were conducted.

@@

lipid/counter-ions  & force field & {[}CaCl$_{2}${]}\,(M)  & \footnote{Number of POPC molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Number of additional cations}N$_{{\rm c}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Reference for simulation files}files\tabularnewline

@@

The headgroup and glycerol backbone C--H bond order parameter magnitudes of 1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine (POPS)
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03~MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15~kHz (R-PDLF experiment) and 5~kHz (S-DROSS experiment).
The following experimental setups were used.

{\emph{C--H bond order parameters from the R-PDLF experiment.}} The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=1.94$~ms and $\tau_2=0.97$~ms.
The used radio frequency pulses had the following nutation frequencies: 46.35~kHz (R18$^7_1$ pulses), 63.45~kHz ($^{13}$C 90$^{\circ}$ and 180$^{\circ}$),
50~kHz (SPINAL64 $^1$H decoupling pulses).
The $t_1$ increment was equal to 10.79~$\mu$s $\times18\times2$, and 32 points in the indirect
dimension were recorded using 1024 scans for each, with a recycle delay of 5~s and a spectral width of 149.5~ppm.

\emph{Order parameter signs from the S-DROSS experiment.}
The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19~ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as $\tau_1=39.4$~$\mu$s and $\tau_2=89.4$~$\mu$s. The used radio frequency pulses had the nutation
frequencies: 63.45~kHz ($^{13}$C 90$^{\circ}$ and 180$^{\circ}$), 50~kHz ($^1$H\,SPINAL64 decoupling).
The $t_1$ increment (dipolar recoupling dimension) was 800~$\mu$s, and a total of 8 points along $t_1$ were
measured using 1024 scans for each, with a recycle delay of 5~s and a spectral width of 149.5~ppm.

@@

\emph{Sample preparation.}
The sample was prepared simply by mixing POPS powder (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
as sodium salt) with water (lipid:water 60:40~wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4~mm NMR rotors.

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis can be conducted using a
Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that utilizes the
MDAnalysis library \cite{agrawal11,gowers16}.
For united atom models, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

The order parameters of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup order parameters at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup order parameters
of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.

Therefore, measuring the PC headgroup order parameters from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

@@

Studies applying the molecular electrometer concept have utilized two different definitions for salt concentration:
the concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids. In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
Nevertheless, in this work we use this former definition to be consistent with the reference
experimental data \cite{roux90}. However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon, and $+0.09$ and $-0.02$
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS ($T=303$~K, $^2$H NMR, 0.1~M of NaCl) \cite{browning80} and 
    POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  We can do this just before submission when citation numbers will not change anymore.}
\end{figure}

@@

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC})
and more accurate force fields are needed
to correctly describe the PC--PS headgroup interactions in MD simulations.
%Having this conclusion in every paragraph gets a bit repetive
